Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-06 ms.tex
Made minor additions to methods section.

\newcommand{\mvir}{M_{\rm vir}}
\newcommand{\rvir}{R_{\rm vir}}
\newcommand{\ncen}{N_{\rm cen}}
\newcommand{\nsat}{N_{\rm sat}}

@@

Important points:
-supersedes previous analyses even for standard HOD.
-cannot rule out assembly bias using galaxy clustering.
-some samples favor assembly bias.

@@

For more than a decade, halo occupation modeling has been used to 
interpret large-scale structure measurements and exploit these measurements 
to constrain galaxy formation models and cosmology \citep[e.g.,][]{yang03,tinker05,zehavi05a,
porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
wake_etal11,yang11a,yang12,leauthaud_etal12,tinker_etal13,cacciato_etal13,
more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions underlying 
halo occupation modeling are: (1) all galaxies reside in dark matter halos that are biased 
tracers of the density field; and (2) galaxies occupy halos as a function of halo masses 
only. It is now well known that halo bias depends upon halo properties other than mass
\citep[e.g.][]{gao_etal05,wechsler06,gao_white07,zentner07,dalal_etal08,lacerna11}, 
an effect called halo assembly bias. If galaxies occupy halos as a function of properties 
other than halo mass, then standard halo occupation methods will be subject to a 
systematic error due to galaxy assembly bias. Several of us have previously shown that 
this error can be significant in an analysis of galaxy clustering and can bias inferences 
about many aspects of galaxy evolution \citep{zentner_etal14}. Consequently, we have 
also developed halo occupation models that enable galaxies to occupy halos in a manner 
that depends upon several halo properties \citep{hearin_etal16}. In this paper, we 
revisit the interpretation of luminosity-dependent galaxy clustering in the Sloan Digital Sky 
Survey (SDSS) Data Release 7 (DR7) data, analyzed previously by \citet{zehavi_etal11}, in the context 
both standard halo occupation models and these new models.

@@

Furthermore, our work demonstrates explicitly that significant assembly bias in 
$M_r$-selected samples from SDSS DR7 cannot be ruled out based on a standard 
analysis of galaxy clustering only. In fact, we find that several samples {\em 
favor galaxy assembly bias to a degree that is statistically significant}. As demonstrated by 
\citet{zentner_etal14}, this conclusion could have important consequences for the 
interpretation of both extant and forthcoming data.

@@

To generate predictions for galaxy clustering, we directly populate dark matter halos with mock galaxies using {\tt Halotools}. In this section we review the ``standard" HOD-style model used in the present work, and in the following section we use the Decorated HOD model described in \citet{hearin_etal16}. In both cases, we will only briefly review the salient features of our methodology here; interested readers can always refer to {\tt halotools.readthedocs.io} for further details.

\subsubsection{Simulation}
All of our analyses are based on the Bolshioi-Planck simulation \citep{riebe_etal11}. 
Bolshoi-Planck was run with cosmological parameters based on \citet{planck13}, $\Omega_{\Lambda} = 0.693, \Omega_{\rm m} = 0.307, \Omega_{\rm b} = 0.048, h = 0.7, {\rm n_s}=0.96$ and $\sigma_8 = 0.82,$ and with a box size of $250$ ${\rm Mpc}/h$ and a particle mass of $m_{\rm p} = 1.35\times10^{8}M_{\odot}/h.$ Further information about the Bolshoi-Planck simulation is available at \url{https://www.cosmosim.org}.

@@

In standard HOD models, there are two types of galaxies: centrals and satellites. The galaxy-halo connection is specified in terms of $P(\ncen|\mvir)$ and $P(\nsat|\mvir),$ the probability that a halo of mass $\mvir$ hosts $\ncen$ central and $\nsat$ satellite galaxies, respectively. $P(\ncen|\mvir)$ is typically a nearest-integer distribution, as a host halo has only either zero or one central galaxy. So occupation statistics of central galaxies are specified by the first moment of $P(\ncen|\mvir)$, which we model as

\begin{equation}
\label{eq:ncen}
\langle N_{\mathrm{cen}} \rangle(\mvir) =
        \frac{1}{2}\left( 1 +
        \mathrm{erf}\left( \frac{\log_{10}\mvir -
        \log_{10}M_{\rm min}}{\sigma_{\log_{10}M}} \right) \right)
\end{equation}
For every host halo in the catalog we draw a random number from a uniform distribution $\mathcal{U}(0, 1);$ for a host halo of present-day virial mass $\mvir,$ a central galaxy is assigned to the halo if the associated random number is less than $\langle N_{\mathrm{cen}} \rangle(\mvir);$ halos with random values exceeding  $\langle N_{\mathrm{cen}} \rangle(\mvir)$ are left devoid of centrals.

We model the distribution $P(\nsat|\mvir)$ to be a Poisson distribution with first moment given by a power law, 
\begin{equation}
\label{eq:nsat}
\langle N_{\rm sat}\rangle(\mvir) = \left( \frac{\mvir - M_0}{M_1} \right)^{\alpha}. 
\end{equation}

@@

We model the velocities of satellite galaxies to be a Gaussian whose first moment is the host halo velocity and whose second moment is the solution to the isotropic Jeans equation for an NFW profile \citep{more09b}, 
\begin{equation}
\sigma^{2}_{r}(\tilde{r}|c) = V_{\rm vir}^{2}\frac{c^{2}\tilde{r}(1 + c\tilde{r})^{2}}{g(c)}\int_{c\tilde{r}}^{\infty}{\rm d}y\frac{g(y)}{y^{3}(1 + y)^{2}}, 
\end{equation}
where $\tilde{r} = r/\rvir,$ $g(x) = \rm{ln}(1+x) - x / (1+x),$ and $V_{\rm vir}^{2} = G\mvir/\rvir.$

@@

Short summary of \citet{hearin_etal16}.

>>>>

\newcommand{\mvir}{M_{\rm vir}}
\newcommand{\rvir}{R_{\rm vir}}
\newcommand{\ncen}{N_{\rm cen}}
\newcommand{\nsat}{N_{\rm sat}}
\newcommand{\ngal}{N_{\rm gal}}
\newcommand{\abias}{\mathcal{A}_{\rm bias}}

@@

Important points:

-supersedes previous analyses even for standard HOD.

-cannot rule out assembly bias using galaxy clustering.

-some samples favor assembly bias.

@@

Halo occupation modeling has frequently 
been used to interpret large-scale structure measurements and exploit these measurements 
to constrain galaxy formation models and cosmology \citep[e.g.,][]{yang03,tinker05,zehavi05a,
porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
wake_etal11,yang11a,yang12,leauthaud_etal12,tinker_etal13,cacciato_etal13,
more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions underlying 
halo occupation modeling are: (1) all galaxies reside in dark matter halos that are biased 
tracers of the density field; and (2) galaxies occupy halos as a function of halo mass 
only. It is now well known that halo bias depends upon halo properties other than mass
\citep[e.g.][]{gao_etal05,wechsler06,gao_white07,zentner07,dalal_etal08,lacerna11}, 
an effect called {\em halo assembly bias}. If galaxies occupy halos as a function of properties 
that affect halo clustering, other than mass, then standard halo occupation methods will be subject to a 
systematic error due to {\em galaxy assembly bias}. Several of us have previously shown that 
this error can be significant in an analysis of galaxy clustering and can bias inferences 
about many aspects of galaxy evolution \citep{zentner_etal14}. Consequently, we have 
developed halo occupation models that enable galaxies to occupy halos in a manner 
that depends upon several halo properties \citep{hearin_etal16}. In this paper, we 
revisit the interpretation of luminosity-dependent galaxy clustering in the Sloan Digital Sky 
Survey (SDSS) Data Release 7 (DR7) data, analyzed previously by \citet{zehavi_etal11}, in the context 
both standard halo occupation models and these new models.

@@

Furthermore, our work demonstrates explicitly that significant assembly bias in 
$M_r$-selected samples from SDSS DR7 cannot be ruled out based on a standard 
analysis of galaxy clustering alone. In fact, we find that the clustering of galaxies in 
several luminosity threshold samples 
{\em favors galaxy assembly bias to a degree that is statistically significant}. 
As suggested by the previous work of \citet{zentner_etal14}, 
this conclusion could have important consequences for the 
interpretation of both extant and forthcoming galaxy survey data.

@@

To generate predictions for galaxy clustering, we directly populate dark matter halos with mock galaxies using {\tt Halotools}. 
We explore halo occupation distribution models (HOD) models in this work 
\citep[e.g.][]{seljak00,ma_fry00,scoccimarro01a,berlind02}, though other 
techniques that can be used to interpret such data, such as the conditional luminosity function 
\citep[CLF, e.g.,][]{yang03,vdBosch13}, exist. In this section, we review the ``standard" HOD 
model used in the present work which assumes that there is no galaxy assembly bias. We refer to 
such a model as ``standard" because all HOD analyses of galaxy clustering to date have assumed no 
galaxy assembly bias. In the following section, we describe 
the Decorated HOD model described in \citet{hearin_etal16}. In both cases, we will only 
review the salient features of our methodology briefly; 
interested readers can always refer to {\tt halotools.readthedocs.io} and 
\citet{hearin_etal16} for further details.

\subsubsection{Simulation}

All of our analyses are based on the Bolshioi-Planck simulation \citep{riebe_etal11}. 
Bolshoi-Planck was run with cosmological parameters based on \citet{planck13}, $\Omega_{\Lambda} = 0.693, \Omega_{\rm m} = 0.307, \Omega_{\rm b} = 0.048, h = 0.7, {\rm n_s}=0.96$ and $\sigma_8 = 0.82,$ within a cubic box 
$250$ ${\rm Mpc}/h$ on a side, requiring a particle mass of $m_{\rm p} = 1.35\times10^{8}M_{\odot}/h.$ 
Further information about the Bolshoi-Planck simulation is available at \url{https://www.cosmosim.org}.

@@

In standard HOD models, central galaxies and satellite galaxies are treated separately, so 
the model is specified by two probability distributions, one for each type of galaxy. 
The galaxy-halo connection is specified in terms of $P(\ncen|\mvir)$ and $P(\nsat|\mvir),$ 
the probability that a halo of mass $\mvir$ hosts $\ncen$ central and $\nsat$ satellite galaxies, 
respectively. $P(\ncen|\mvir)$ is typically a nearest-integer distribution, as a host halo has only 
either zero or one central galaxy. Consequently, the occupation statistics of central galaxies are 
specified by the first moment of $P(\ncen|\mvir)$, which we model as 
%
\begin{equation}
\label{eq:ncen}
\langle N_{\mathrm{cen}} \rangle(\mvir) =
        \frac{1}{2}\left( 1 +
        \mathrm{erf}\left( \frac{\log (\mvir) -
        \log (M_{\rm min})}{\sigma_{\log M}} \right) \right)
\end{equation}
%
For every host halo in the catalog we draw a random number from a uniform distribution $\mathcal{U}(0, 1);$ for a host halo of present-day virial mass $\mvir,$ a central galaxy is assigned to the halo if the associated random number is less than $\langle N_{\mathrm{cen}} \rangle(\mvir);$ halos with random values exceeding  $\langle N_{\mathrm{cen}} \rangle(\mvir)$ are left devoid of centrals. 
The parameter $\log (M_{\rm min})$ specifies the halo mass at which the halo has a 50\% probability of hosting a 
central galaxy, while the parameter $\sigma_{\log M}$ specifies the rate at which $\langle N_{\mathrm{cen}} \rangle (\mvir)$ 
transitions from zero to unity, with smaller values of $\sigma_{\log M}$ corresponding to a more rapid transition.

We model the distribution $P(\nsat|\mvir)$ as a Poisson distribution with first moment given by a power law, 
%
\begin{equation}
\label{eq:nsat}
\langle N_{\rm sat}\rangle(\mvir) = \left( \frac{\mvir - M_0}{M_1} \right)^{\alpha}. 
\end{equation}
%
The parameter $M_0$ allows the power-law to be truncated more rapidly at low masses and we 
set $\langle N_{\rm sat}\rangle (\mvir) = 0$ for $\mvir < M_0$.

The five parameters of the standard HOD models are varied in standard analyses are $\log (M_{\rm min})$, 
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$, though, as we show below, central galaxies usually 
outnumber satellite galaxies in by a factor of several, so $\log (M_{\rm min})$ and $\sigma_{\log M}$ usually 
vary along a narrow degeneracy that fixes the total galaxy number density to the observed value. There are 
many particular choices that can be made for the functional forms of $\langle N_{\rm cen}\rangle (\mvir)$ and 
$\langle N_{\rm sat}\rangle (\mvir)$. We have made choices that mimic the standard SDSS DR7 analysis of 
\citet{zehavi_etal11}, to expedite comparisons with their results.

@@

We model the velocities of satellite galaxies as a Gaussian whose 
first moment is the host halo velocity and whose second moment is the 
solution to the isotropic Jeans equation for an NFW profile \citep{more09b}, 
\begin{equation}
\sigma^{2}_{r}(\tilde{r}|c) = V_{\rm vir}^{2}\frac{c^{2}\tilde{r}(1 + c\tilde{r})^{2}}{g(c)}\int_{c\tilde{r}}^{\infty}{\rm d}y\frac{g(y)}{y^{3}(1 + y)^{2}}, 
\end{equation}
where $\tilde{r} = r/\rvir,$ $g(x) = \rm{ln}(1+x) - x / (1+x),$ and $V_{\rm vir}^{2} = G\mvir/\rvir.$ 
In this way, we model the influence of redshift-space distortions on our predicted 
two-point functions.

@@

In addition to the standard occupation statistics described in the previous section, 
in this paper we also use the decorated HOD formalism to connect galaxies t
o dark matter halos in a manner that has simultaneous dependence on both $\mvir$ and halo concentration. 
Briefly, we use Equations \ref{eq:ncen} and \ref{eq:nsat} as our ``baseline" first occupation moments. 
At fixed $\mvir,$ halos are divided into one of two categories, those of high- and low-concentration, 
depending on whether the concentration of the halo places it above or below the rank-order percentile 
$f_{\rm split},$ which we keep fixed to $f_{\rm split}=0.5$ throughout the paper for simplicity. 
High-concentration halos have a different first occupation moment relative to low-concentration 
halos of the same mass, $\langle\ngal|\mvir, c_{\rm high}\rangle \neq \langle\ngal|\mvir, c_{\rm low}\rangle.$ The difference between the first moment of high- and low-concentration halos is modulated by $\abias,$ the novel parameter of the decorated HOD governing assembly bias. Values of $\abias=\pm1$ correspond to the maximum strength of assembly bias allowable by the constraint that the model preserves the marginalized first moment, $\langle\ngal|\mvir\rangle;$ thus regardless of the value of $\abias,$ in the decorated HOD the marginalized first moment of centrals and satellites are {\em unchanged from the baseline value defined by Equations \ref{eq:ncen} and \ref{eq:nsat}}. 
Decorated HOD models all have the same HODs, when averaged over all halos at fixed mass, as standard HOD models. The only 
change in Decorated HOD models is whether or not an additional property also modulates halo occupation at fixed halo mass. A 
value of $\abias = 0$ indicates no galaxy assembly bias whatsoever. We refer the reader to \citet{hearin_etal16} for 
further details about the decorated HOD.

In the present work, we fix our model to the simplest class of galaxy assembly bias models, though a recipe for 
generalizing to more complicated models can be found in \citet{hearin_etal16}. In particular, we split halos into 
two populations as specified in the previous paragraph. We then populate halos with satellite galaxies specified 
by an assembly bias parameter $-1 \le A_{\rm sat} \le 1$ and central galaxies with a distinct assembly bias 
parameter $-1 \le A_{\rm cen} \le 1$. These two additional parameters are varied in our analyses that include 
assembly bias, making the number of parameters that vary in these analyses seven. As we show below, 
these additional parameters are in many instances poorly constrained by clustering data of the quality of 
SDSS DR7 alone, so exploring more complex models of assembly bias does not yet seem justified.
